---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# peruse

<!-- badges: start -->
<!-- badges: end -->

The {peruse} package is aimed at making it easier to generate irregular
sequences that are difficult to generate with existing tools. The heart
of {peruse} is the `S3` class `Iterator`. An `Iterator` allows the
user to write an arbitrary R expression that returns the next
element of a sequence of R objects. It then saves the state of the `Iterator`, meaning the next time `yield_next` is called, the subsequent element of the sequence will be returned. The `Iterator` also has helper functions `yield_more()`, `move_next()`, `move_more()`, and `yield_while()` that provide variations on `yield_next()`.

The package also provides a simple, tidy API for set building, allowing the user to generate a set consisting of the elements of a vector that meet specific criteria. This can either return a vector consisting of all the chosen elements or it can return an `Iterator` that lazily generates the chosen elements.

## Installation

You can install the released version of peruse from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("peruse")
```


And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("jacgoldsm/peruse")
```
## Example
### Collatz Sequence

A Collatz sequence is a particular sequence of natural numbers that mathematicians think always reaches `1` at some point, no matter the starting point. We can't prove that one way or the other, but we can create an `Iterator` that lazily generates a Collatz sequence until it reaches `1`:

```{r example}
library(peruse)
expr <- "if (n %% 2 == 0) n <- n / 2 else n <- n*3 + 1"
  
# Collatz generator starting at 50
collatz <- Iterator(result = expr,
                    initial = c(n = 50),
                    yield = n)

yield_while(collatz, cond = "n != 1L")
```
### Random Walk with Drift
Random Walks, with or without drift, are one of the most commonly used type of stochastic processes. How can we simulate one with {peruse}?

```{r}
set.seed(1)
expr <- 'n <- n + sample(c(-1L, 1L), size = 1L, prob = c(0.25, 0.75))'
rwd <- Iterator(result = expr,
                initial = c(n = 0),
                yield = n)


Value <- yield_while(rwd, "n != 50L & n != -50L")

plot(Value, main = "The Value of the Iterator after a Given Number of Iterations")
```

Here, we can see that `seq` gets to `50` after about `100` iterations when it is weighted `3:1` odds in favor of adding `1` over adding `-1` to the prior value.

### Primes
How about generating all the prime numbers between `1` and `100`? We can easily do that with the set-builder API:
```{r}
2:100 %>% that_for_all(range(2, .x)) %>% we_have(~.x %% .y != 0)
```
But how about if we want to generate the first 100 prime numbers? We don't know the range of values this should fall in (well, mathematicians do), so we can use laziness to our advantage:
```{r}
primes <- 2:10000 %>%
            that_for_all(range(2, .x)) %>% 
            we_have(~.x %% .y != 0, "Iterator")

sequence <- yield_more(primes, 100)

sequence
```
